Stochastic blockmodels for relational events

### 2012-03-14 docs/slides/slides.tex
Sunbelt talk submitted to Carter.

\setlength{\unitlength}{\textwidth} 
\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
\title{Stochastic blockmodels for relational events}
% \subtitle{}
\author[DuBois, Christopher] % (optional, for multiple authors)
{Christopher~DuBois\inst{1} \and Carter~Butts\inst{2} \and Padhraic~Smyth\inst{3}}
\institute[] % (optional)
{
  \inst{1}%
Department of Statistics \\
University of California, Irvine
  \and
  \inst{2}%
Department of Sociology \\
University of California, Irvine
  \and
  \inst{2}%
Department of Computer Science \\
University of California, Irvine
}
\date[Sunbelt 2012] % (optional)
{Sunbelt 2012}
\begin{document}
\frame{\titlepage}

@@

\begin{frame}{Relational event models: Modeling the \emph{hazard}}
Multiplicative models for the hazard of a dyad $(i,j)$:
$$\log \lambda_{ij}(t | \cdot) = \boldsymbol{\beta}' \mathbf{s}(t,i,j)$$
\begin{itemize}
\item $\lambda_{ij}(t | \cdot)$: ``rate'' of interaction from $i$ to
  $j$ at time $t$
\item $\boldsymbol{\beta}$: vector of model parameters
\item $\mathbf{s}(t,i,j)$: vector of sufficient statistics for $(i,j)$
  given the events before time $t$
\end{itemize}
\end{frame}

@@

\vspace{-.8cm}
\begin{align*}
\uncover<5->{
\mbox{Mixing rates} & & \boldsymbol{\eta} =& \left[
\begin {array}{cc}
 \eta_{11}& \eta_{12} \\
 \eta_{21}& \eta_{22} 
\end {array}
\right]
= \left[
\begin {array}{cc}
.8 & .2\\.2 & .8
\end {array}
\right] \\
}
\uncover<6->{
\mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi) \\
}
\uncover<7->{
\mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
}
\end{align*}

@@

\textbf{Degree effects} count the number of previous events
\begin{itemize}
\item sent by sender
\item received by sender
\item sent be receiver
\item received by receiver
\item previous occurrences
\end{itemize}

@@

\begin{frame}{Specifying degree effects (and explosion)}

@@

\begin{frame}{Synthetic example: slice sampling for $\beta$ and Gibbs on
    $\mathbf{z}$}
\begin{picture}(0.0,0.0)
   \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-estimates}}
   \put(.65,-0.25){\includegraphics[width=0.55\textwidth]{../../figs/synthetic/counts}}
\end{picture}

@@

\begin{frame}{Application: Dyadic interactions on Twitter}
  \begin{itemize}
  \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 with hashtag \texttt{\#rstats}.
  \item Selecting tweets beginning with the \texttt{@} symbol, and mark the first mentioned user as the recipient.
  \item Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
  \item We consider a subset 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
  \end{itemize}
\end{frame}

\begin{frame}{Application: Dyadic interactions on Twitter}
Prediction experiments and model comparison
\end{frame}

\begin{frame}{Application: Dyadic interactions on Twitter}
Plot parameter estimates for winning model
Compare to fitting a single relational event model
\end{frame}

\begin{frame}{Model adequacy}
Plot parameter estimates for winning model
Compare to fitting a single relational event model
\end{frame}

\begin{frame}{Discussion}
Interpretation of parameter estimates
  \begin{itemize}
  \item Method clusters with respect to the relational event dynamics for a set of actors
  \end{itemize}

Main takeaways
\begin{itemize}
\item Detailed, highly dependent model for local structure\\ (using relational event models)
\item Latent variable model to capture meso-level structure \\(using stochastic block model)
\end{itemize}
\end{frame}

\begin{frame}{Future directions}
  \begin{itemize}
  \item Benefits of mixed-membership ideas
  \item Dirichlet process instead of fixed $K$ (with split-merge sampling?)
  \end{itemize}
\end{frame}

\begin{frame}{References}

>>>>

\setlength{\unitlength}{\textwidth} 
\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
\title{Stochastic blockmodels for relational events}
% \subtitle{}
\author[DuBois, Christopher] % (optional, for multiple authors)
{Christopher~DuBois\inst{1} \and Carter~Butts\inst{2} \and Padhraic~Smyth\inst{3}}
\institute[] % (optional)
{
  \inst{1}%
Department of Statistics \\
University of California, Irvine
  \and
  \inst{2}%
Department of Sociology \\
University of California, Irvine
  \and
  \inst{2}%
Department of Computer Science \\
University of California, Irvine
}
\date[Sunbelt 2012] % (optional)
{Sunbelt 2012}

\renewcommand\footnotemark{}
\renewcommand*\footnoterule{}

\begin{document}
\frame{\titlepage}

@@

\begin{frame}{Relational event models}
Multiplicative models for the hazard of a dyad $(i,j)$:
$$\log \lambda_{ij}(t | \cdot) = \boldsymbol{\beta}' \mathbf{s}(t,i,j)$$
\begin{itemize}
\item $\lambda_{ij}(t | \cdot)$: ``rate'' of interaction from $i$ to
  $j$ at time $t$
\item $\boldsymbol{\beta}$: vector of model parameters
\item $\mathbf{s}(t,i,j)$: vector of sufficient statistics for $(i,j)$
  given the events before time $t$
\end{itemize}
\vspace{2cm}
\footnotesize{See \cite{Butts2008, Vu2011,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011}.}
\end{frame}

@@

\vspace{-.8cm}
\begin{align*}
\uncover<5->{
\mbox{Mixing rates} & & \boldsymbol{\eta} =& \left[
\begin {array}{cc}
 \eta_{11}& \eta_{12} \\
 \eta_{21}& \eta_{22} 
\end {array}
\right]
= \left[
\begin {array}{cc}
.8 & .2\\.2 & .8
\end {array}
\right] \\
}
\uncover<6->{
\mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi) \\
}
\uncover<7->{
\mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
}
\end{align*}
\uncover<8->{
\footnotesize{See \cite{Nowicki2001, Kemp}.}
}
\end{frame}

@@

\textbf{Degree effects} count the number of previous events
\begin{itemize}
\item sent by sender
\item received by sender
\item sent be receiver
\item received by receiver
\item previous occurrences of this dyad
\end{itemize}

@@

% \begin{frame}{Specifying degree effects (and explosion)}

% \end{frame}

@@

\begin{frame}{Synthetic example: slice sampling for $\beta$ and Gibbs on
    $\mathbf{z}$}
\begin{picture}(0.0,0.0)
   \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-estimates}}
   \put(.55,-0.25){\includegraphics[width=0.55\textwidth]{../../figs/synthetic/counts}}
\end{picture}

@@

\begin{frame}{Prediction experiment: Synthetic data}
Mean log likelihood on observed events and held out events.
\input{../../figs/synthetic/results.tex}
\end{frame}

\begin{frame}{Application: Modeling email (88 people, 2000 events)}

\begin{picture}(0.0,0.0)
   \put(-.05,-0.25){\includegraphics[scale=.4]{../../figs/eckmann-small/countsa}}
 \uncover<2->{  \put(.5,-0.25){\includegraphics[scale=.5]{../../figs/eckmann-small/params-estimates}}}
\end{picture}
\end{frame}

\begin{frame}{Prediction experiment: Eckmann subset}
Mean log likelihood on observed events and held out events.
\input{../../figs/eckmann-small/results.tex}
\end{frame}

% \begin{frame}{Application: Dyadic interactions on Twitter}
%   \begin{itemize}
%   \item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 with hashtag \texttt{\#rstats}.
%   \item Selecting tweets beginning with the \texttt{@} symbol, and mark the first mentioned user as the recipient.
%   \item Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
%   \item We consider a subset 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
%   \end{itemize}
% \end{frame}

% \begin{frame}{Application: Dyadic interactions on Twitter}
% Prediction experiments and model comparison
% \end{frame}

% \begin{frame}{Application: Dyadic interactions on Twitter}
% Plot parameter estimates for winning model
% Compare to fitting a single relational event model
% \end{frame}

% \begin{frame}{Model adequacy}
% Plot parameter estimates for winning model
% Compare to fitting a single relational event model
% \end{frame}

\begin{frame}{Discussion}
Interpretation of parameter estimates
  \begin{itemize}
  \item The proposed method clusters individuals with respect to
    shared relational event dynamics with other clusters
  \end{itemize}

\uncover<2->{
Main takeaways
\begin{itemize}
\item Detailed, highly dependent model for local structure\\ (using relational event models)
\item Latent variable model to capture meso-level structure \\(using stochastic block model)
\end{itemize}
}

\uncover<3->{
Future directions
  \begin{itemize}
  \item Consider a mixed-membership approach
  \item Dirichlet process instead of fixed $K$
  \end{itemize}
}
\end{frame}

\begin{frame}[allowframebreaks]{References}
\tiny
\bibliographystyle{apalike}
\bibliography{../refs}
\end{frame}

\begin{frame}{Thanks}
Work supported by National Defense Science and Engineering Graduate
Fellowship.

\vspace{2cm}
Special thanks to Carter Butts and Padhraic Smyth.
